---
title:  "Constructing indexes of abundance"
subtitle: "Fish 576, Week 2"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css", "slides-style.css"]
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(message=FALSE, fig.retina=3)
library(ggplot2)
theme_set(theme_classic(base_size = 16))
options(htmltools.dir.version = FALSE)
```

```{r xaringan-tile-view, echo=FALSE, purl=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()

```

---

## What is an index?

* Indexes of abundance give information on relative changes in population abundance or biomass over time, especially when coupled with information on the lengths and ages that index tends to capture.

--

* **Fishery independent** indexes generally have consistent and well-documented effort, easy to calculate CPUE
  * Used for assessments in this class
  
--

* **Fishery dependent** indexes standardize fishery catch rates by information on effort from monitoring programs, logboks, etc.

---
class: center, middle

# Index evolution:

<br>

### Design based ➜ DeltaGLMM ➜ Geospatial models

---

## Design-based indexes

* Essentially a weighted average

--

* Stratification: partitioning the population so that samples within a **stratum** are more homogeneous than samples across strata.
  * Decreases sampling variance.
  
--

* What variables might we stratify by?

--

  * Possible ideas: Depth, state, habitat type, latitude (note continuous variables must be binned)

---

## Calculation for a single year

* Calculate CPUE for tow $i$ which sweeps over area $a_i$

$$CPUE_i = C_i/a_i$$

--

* Calculate average CPUE in stratum $j$ which has $n_j$ tows

$$ \overline{CPUE}_j = \tfrac{1}{n_j} \sum_i^{n_j} CPUE_i$$

--

* Expand to total estimated biomass in stratum $i$, which has area $A_j$

$$\hat b_j = \overline{CPUE}_j \cdot A_j$$

---

## Calculation, continued

* Sum over all strata to obtain index estimate

$$\hat B = \sum_j \hat b_j$$

---

## Design-based index in R

```{r, eval=FALSE}
remotes::install_github("pfmc-assessments/nwfscSurvey")
```

```{r plot_cpue, fig.show='hide'}
library(nwfscSurvey)
# Pull survey data from NWFSC data warehouse
catch <- PullCatch.fn(Name = "rex sole", # all lowercase 
                      SurveyName = "NWFSC.Combo",
                      SaveFile = FALSE)
plot_cpue(dir = NULL, catch)
```

---

```{r, ref.label='plot_cpue', echo=FALSE, purl=FALSE, fig.show='hold', out.width='50%'}
```

---

## Defining strata

```{r strata, results='hide'}
strata <- CreateStrataDF.fn(
  names = c("shallow_s", "mid_s", "deep_s", "shallow_n", "mid_n", "deep_n"), 
  depths.shallow = c( 55,   200, 300,    55, 200, 300),
  depths.deep    = c(200,   300, 400,   200, 300, 400),
  lats.south     = c( 32,    32,  32,    42,  42,  42),
  lats.north     = c( 42,    42,  42,    49,  49,  49))
strata
```
--
```{r, ref.label='strata', echo=FALSE, comment=NA, purl=FALSE}

```

---

## Calculating index

```{r index_plot, fig.show='hide'}
biomass <- Biomass.fn(dir = NULL, 
                      dat = catch,  
                      strat.df = strata)
ggplot(biomass$Bio) +
  geom_pointrange(aes(x = as.numeric(Year),
                      y = Value,
                      ymin = exp(log(Value) - 1.96*seLogB),
                      ymax = exp(log(Value) + 1.96*seLogB))) +
  labs(x = 'Year', y = 'Index')
```

---

```{r, ref.label='index_plot', echo=FALSE, purl=FALSE}

```

---

## Exercise

Calculate design-based index for a survey for one of the two species. Options are:

* Shortspine thornyhead
  * Triennial
  * AFSC slope survey
  * NWFSC slope survey
  * NWFSC trawl/WCGBTS/combined survey
  * Rex Sole
  * Triennial
* Rex Sole
  * Triennial
  
---

## Problems with design-based index

* High year-to-year variance for species that cluster or school, like many rockfish

* Hard to account for auxiliary information such as vessel effects

* Does not handle unbalanced sampling designs well

* Observations near each other should be more similar, information sharing could be used to decrease variance

---

## DeltaGLMM

* Encounter model

$$\Pr(C_{i,j,y,v}>0) = \text{logit}^{-1}(Y_{p,y} + S_{p,j} + \gamma_{p,v}), 
\gamma_{p,v} \sim N(0,\sigma_p)$$

--

* Positive catch rate model

$$\text{E}[C_{i,j,y,v} | C_{i,j,y,v}>0] = \exp(Y_{r,y} + S_{r,j} + \gamma_{r,v}) \\
\gamma_{r,v} \sim N(0,\sigma_r)$$

* $C_{i,j,y,v} | C_{i,j,y,v} > 0$ follows continuous positive probability distribution (e.g., lognormal, gamma)

--

* $y$ is year, $v$ is vessel, $Y$, $S$, $\gamma$ are year, stratum, and vessel effects, respectively

---

## DeltaGLMM index

$$\hat B_y = \sum_j A_j \text{logit}^{-1}(Y_{p,y} + S_{p,j}) \cdot
\exp(Y_{r,y} + S_{r,j})$$

--

* Note that we dropped the vessel effect

--

* This is a simple example. What else could you include?

--

  * Year-stratum interactions, pass (early/late), ?
  
---

## Geostatistical models

* deltaGLMMs solve many problems with design-based indexes, but do not permit information sharing among locations close to each other in space and time

--

* Geostatistical encounter model

$$\Pr(C_{i,j,y,v,s}>0) = \text{logit}^{-1}(Y_{p,y} + S_{p,j} + \gamma_{p,v} +
\omega_{p,s} + \psi_{p,s,y}) \\ 
\gamma_{p,v} \sim N(0,\sigma_p) \\
\bar{\omega}_p \sim \text{MVN}(0, \Sigma_{p,\omega}), 
\bar{\psi}_{p,y} \sim \text{MVN}(0, \Sigma_{p,\psi})$$

--

* Positive catch rate model adjusted similarly

---

## Geostatistical index

$$\begin{align}
\hat{B}_y =  \int_s & \text{logit}^{-1}(Y_{p,y} + S_{p,j} +
  \omega_{p,s} + \psi_{p,s,y}) \cdot \\
  & \exp(Y_{r,y} + S_{r,j} + \omega_{r,s} + \psi_{r,s,y}) ds
\end{align}$$

--

$$\begin{align}
\approx \sum_{s_k \in \text{grid}} & \left[ \text{logit}^{-1}(Y_{p,y} + S_{p,j} +
  \omega_{p,s} + \psi_{p,s,y}) \cdot \right. \\
  & \left. \exp(Y_{r,y} + S_{r,j} + \omega_{r,s} + \psi_{r,s,y}) \right] 
\end{align}$$

--

* Kelli Johnson at NWFSC will be producing all fishery independent indexes for PFMC assessments this year using {sdmTMB}, which fits geostatistical models