---
title:  "Estimating growth"
subtitle: "Fish 576, Week 3"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["xaringan-themer.css", "slides-style.css"]
    lib_dir: libs
    nature:
      ratio: '4:3'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc( (%current% - 1) / (%total% - 1) * 100%);">
          </div>
        </div>`
---
layout: true

.footnote[U.S. Department of Commerce | National Oceanic and Atmospheric Administration | National Marine Fisheries Service]


```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.retina=3)
options(htmltools.dir.version = FALSE)
```

```{r, include=FALSE}
library(ggplot2)
library(magrittr)
theme_set(theme_classic(base_size = 16))
```


```{r xaringan-tile-view, echo=FALSE, purl=FALSE}
# this gives you a tile navigation if you type "O" at any time
xaringanExtra::use_tile_view()

```

---

## Why estimate growth?

Ageing fish takes time. Measuring fish is fast.

---

## Von Bertalanffy growth function

$$\tfrac{dL}{dt} = k(L_\infty - L)$$
--

* $k$ is the "Brody growth coefficient"

* $L_\infty$ is the asymptotic size as the individual gets old

--

* The solution to this differential equation, up to a constant, is:

$$L(t) = L_\infty\left[1-e^{-k(t-t_0)}\right]$$

* $t_0$ is the x-intercept.

---

Let's look at data for canary rockfish

```{r, eval=FALSE}
library(nwfscSurvey)
canary.bio <- PullBio.fn(Name = "canary rockfish", 
                         SurveyName = "NWFSC.Combo", 
                         SaveFile = FALSE)
```

```{r, include=FALSE, purl=FALSE}
library(here)
load(here('code/Bio_All_NWFSC.Combo_2023-01-20.rda'))
canary.bio <- Data
```

```{r growth_scatter, fig.show='hide'}
ggplot(canary.bio) +
  geom_point(aes(x = Age, y = Length_cm, col = Sex), 
             alpha = 0.25) 
```

---

```{r, ref.label='growth_scatter', echo=FALSE, purl=FALSE}

```

---

## What to do with unsexed fish?

--

```{r}
unsexed.index <- which(canary.bio$Sex=='U' & 
                         !is.na(canary.bio$Age))
set.seed(20398)
males <- sample(unsexed.index, 
                size = floor(length(unsexed.index)/2), 
                replace = FALSE)
canary.bio2 <- canary.bio
canary.bio2$Sex[males] <- 'M'
canary.bio2$Sex[setdiff(unsexed.index, males)] <- 'F'
```

---

## Fit the VBGF using `nls()`. 

```{r}
canary.bio2$Sex <- factor(canary.bio2$Sex)

vbgf.nls <- nls(Length_cm ~ linf[Sex]*
                  (1-exp(-k[Sex]*(Age-t0[Sex]))), 
                data = canary.bio2, 
                start = list(linf = rep(55,2), 
                             k = rep(0.2,2), 
                             t0 = rep(0,2)))
```

---

```{r}
summary(vbgf.nls)
```

---

```{r}
summary(vbgf.nls, cor=TRUE)$correlation %>%
  round(2)
```

--

Correlation between $k$ and $L_\infty$ is very high!

---

## Schnute reparameterization

Choose ages $a_1$ and $a_2$ that bracket the well-sampled range of ages. Then:

$$L(t) = L_{a_1} + (L_{a_2}-L_{a_1})\frac{1-e^{-k(t-a_1)}}{1-e^{-k(a_2-a_1)}}$$
--

* $k$ defined as previously

* $L_{a_1}$ and $L_{a_2}$ are estimated instead of $L_\infty$ and $t_0$

---

Choose $a_1=1$, $a_2=25$.

```{r}
vbgf.nls2 <- nls(Length_cm ~ la1[Sex] + (la2[Sex] - la1[Sex]) * 
                   (1-exp(-k[Sex]*(Age-1))) / 
                   (1-exp(-k[Sex]*24)), 
                 data = canary.bio2, 
                 start = list(la1 = rep(12,2), 
                              la2 = rep(50,2), 
                              k = rep(0.2,2)))
```

---

```{r}
summary(vbgf.nls2)
```

---

```{r}
summary(vbgf.nls2, cor=TRUE)$correlation %>%
  round(2)
```

--

Somewhat better

--

Note this is the parameterization stock synthesis uses. However, SS3 will assume linear growth between age 0 and $a_1$.

---

## Residuals 

.pull-left-wide[
```{r, echo=FALSE}
plot(canary.bio2$Age[!is.na(canary.bio2$Age)], resid(vbgf.nls2),
     xlab = 'Age')
abline(h=0)
```
]

--

.pull-right-narrow[
There is some fanning, especially between ages 0-10. It can be more pronounced sometimes.
]

---

## Lognormal errors

`nls()` assumes normal errors. We could try fitting 
</br>
`nls(log(L) ~ log(VBGF))`.

What is wrong with this?

--

You will be fitting to the median length, not the mean. We need a log-normal bias correction. 

Recall, if $\log(X)\sim N(\mu, \sigma^2)$, then
$$E[X] = \exp\left(\mu + \tfrac{\sigma^2}{2}\right)$$

To fit log normal errors to the *mean* length, we will write our own likelihood function.

---

## Lognormal errors

```{r}
vbgf.loglik <- function(log.pars, dat.m, dat.f, a1, a2) {
  pars <- exp(log.pars)
  
  l.pred.m <- pars['la1.m'] + (pars['la2.m'] - pars['la1.m']) * 
    (1-exp(-pars['k.m']*(dat.m$Age - a1))) / 
    (1-exp(-pars['k.m']*(a2-a1)))
  
  l.pred.f <- pars['la1.f'] + (pars['la2.f'] - pars['la1.f']) * 
    (1-exp(-pars['k.f']*(dat.f$Age - a1))) / 
    (1-exp(-pars['k.f']*(a2-a1)))
  
  nll <- -dlnorm(x = c(dat.m$Length_cm, dat.f$Length_cm), 
                 meanlog = log(c(l.pred.m, l.pred.f)) -
                   pars['cv']^2/2, 
                 sdlog = pars['cv'], log = TRUE) %>%
    sum()
  return(nll)
}

```

---

## Fit using `optim()`
```{r}
dat.m <- dplyr::filter(canary.bio2, Sex == 'M', !is.na(Age))
dat.f <- dplyr::filter(canary.bio2, Sex == 'F', !is.na(Age))

pars.init <- log(c(la1.m = 12, la1.f = 12, 
                   la2.m = 50, la2.f = 55, 
                   k.m = .17, k.f = .14, cv = .1))

vbgf.optim <- optim(pars.init, vbgf.loglik, 
                    dat.m = dat.m, dat.f = dat.f, 
                    a1 = 1, a2 = 25)

```

---
```{r, results='hold'}
exp(vbgf.optim$par)[1:4]
exp(vbgf.optim$par)[5:7]
```
--

Can calculate SEs of *log*-estimates using option `hessian = TRUE`. 

For SEs of model parameters, use delta method or MCMC.

---
# Compare fits
```{r, echo=FALSE}
nls.f <- coef(vbgf.nls2)[c(1,3,5)]
nls.m <- coef(vbgf.nls2)[c(2,4,6)]
optim.f <- exp(vbgf.optim$par)[c(2,4,6)]
optim.m <- exp(vbgf.optim$par)[c(1,3,5)]

ggplot(canary.bio2) +
  geom_point(aes(x = Age, y = Length_cm, col = Sex), alpha = 0.25) +
  geom_line(aes(x = Age, y = nls.f[1] + (nls.f[2] - nls.f[1]) *
                  (1-exp(-nls.f[3]*(Age-1))) / 
                  (1-exp(-nls.f[3]*24)), linetype='1'), col = 'blue') +
  geom_line(aes(x = Age, y = nls.m[1] + (nls.m[2] - nls.m[1]) *
                  (1-exp(-nls.m[3]*(Age-1))) / 
                  (1-exp(-nls.m[3]*24)), linetype='1'), col = 'red') +
  geom_line(aes(x = Age, y = optim.f[1] + (optim.f[2] - optim.f[1]) *
                  (1-exp(-optim.f[3]*(Age-1))) / 
                  (1-exp(-optim.f[3]*24)), linetype='2'), col = 'blue') +
  geom_line(aes(x = Age, y = optim.m[1] + (optim.m[2] - optim.m[1]) *
                  (1-exp(-optim.m[3]*(Age-1))) / 
                  (1-exp(-optim.m[3]*24)), linetype='2'), col = 'red') +
  scale_color_manual(values = c('F' = 'blue', 'M' = 'red')) +
  # scale_linetype_identity(name = "Model fit",
  #                         breaks = c("solid", "dotted"),
  #                         labels = c("normal", "lognormal"),
  #                         guide = "legend") +
  NULL
```


